---
title: "A Young Man's Game"
author: "Michael Foley"
date: "`r Sys.Date()`"
output: 
  html_document:
    css: "style.css"
    theme: flatly
    toc: true
    toc_float: true
    highlight: haddock
    code_folding: show
knit: (function(inputFile, encoding) {
  rmarkdown::render(inputFile, encoding = encoding, output_dir = "../docs") })
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE
)
```

A major league baseball career can expect to last into the player's mid-thirties. Advances in sports training might extend careers. Using a data set of every person who has played in a Major League game, this analysis investigates changes in career longevity.

This document is the detailed work log. The final report is published here, and a summarized account is in the repository [ReadMe](../README.md) and blog.

## Setup

In addition to the usual packages, include **survival** for modeling, **survminer** for visualization, and **gtsummary** for summaries.

```{r warning=FALSE, message=FALSE}
library(tidyverse)
library(janitor)
library(glue)
library(scales)
library(lubridate)
library(extrafont)
library(survival)
library(survminer)
library(gtsummary)

# More color palettes here:
# https://www.schemecolor.com/tag/mlb
mlb_palette <- c("navy_blue" = "#0C2340", "red" = "#E31937")

mlb_theme <- theme(
  text = element_text(size = 16, family = "TR Plaza", color = mlb_palette["navy_blue"]),
  panel.grid.minor = element_blank()
)
```

## Data Management

Retrosheet publishes biographical information of every person who has ever played Major League baseball ([codebook and file](https://www.retrosheet.org/biofile.htm)). I downloaded the file to my data dir.

```{r}
# https://www.retrosheet.org/biofile.htm
bio_0 <- readr::read_csv(
  file.path("../data/BIOFILE.TXT"), 
  # Initially treat all cols as character; transform later.
  col_types = paste(rep("c", 33), collapse = "")
)
```

The file consists of `r nrow(bio_0) %>% comma(1)` rows and `r ncol(bio_0)` columns, one row per player, manager/coach, and umpire. The column definitions are in the [codebook](https://www.retrosheet.org/biofile.htm). This analysis is limited to players, so I'll filter on non-null `PLAY DEBUT`. I'll drop the non-player cols and irrelevant/unused cols. I'll also use snake-case for the names.

```{r}
bio_1 <- bio_0 %>%
  filter(!is.na(`PLAY DEBUT`)) %>%
  select(-c(`MGR DEBUT`:`UMP LASTGAME`, `NAME CHG`, `BAT CHG`, `BIRTH NAME`)) %>% 
  clean_names(case = "snake")
```

That reduces the data set to `r nrow(bio_1) %>% comma(1)` x `r ncol(bio_1)`. While Working through the time length calculations I discovered a few data irregularities.

```{r}
bio_2 <- bio_1 %>%
  mutate(
    # Birth dates with unknown day have format MM/ /YYYY, MM /YYYY, or MM/YYYY. 
    # Impute mid-month.
    birthdate = str_remove_all(birthdate, " "),
    birthdate = str_replace(birthdate, "\\/\\/", "\\/15\\/"),
    birthdate = if_else(
      str_count(birthdate, "\\/") == 1,
      str_replace(birthdate, "\\/", "\\/15\\/"),
      birthdate
    ),
    # Birth dates with unknown month are in format YYYY. Impute mid-year.
    birthdate = if_else(!str_detect(birthdate, "\\/"), 
                        paste0("06/30/", birthdate), birthdate),
    # Same with death dates.
    deathdate = if_else(!str_detect(deathdate, "\\/"), 
                        paste0("06/30/", deathdate), deathdate),
    # Dan McGarvey has a typo!
    birthdate = if_else(birthdate == "12/2/2887", "12/2/1887", birthdate),
    # So does Eddie Quick.
    birthdate = if_else(birthdate == "12/1527/1881", "12/15/1881", birthdate),
    # Andrew Carignan was not born in 1868!
    birthdate = if_else(playerid == "caria001", "07/23/1986", birthdate),
    # Melvin Dorta is not dead.
    deathdate = if_else(playerid == "dortm001", NA_character_, deathdate),
    # But Mike Marshall is.
    deathdate = if_else(playerid == "marsm101", "05/31/2021", deathdate),
    across(contains(c("date", "debut", "lastgame")), mdy),
    weight = as.numeric(weight)
  )
```

Next I'll engineer some time length cols to support survival models. Three good ways to look at longevity are player lifetimes, career lengths, and age at career-end. 

```{r}
# Last file update is last known observation date.
bio_last_updated <- ymd("2021-12-02")

# Need to be careful with missing birth dates, 
# and null death dates (which may be incorrectly inferred as living).
bio_3 <- bio_2 %>%
  mutate(
    life_yrs = time_length(
      interval(birthdate, coalesce(deathdate, bio_last_updated)), 
      "years"
    ),
    # If age > 110 years, or there is no death date but there is another 
    # indicator of death, then you do not know how long they lived.
    life_yrs = if_else(
      life_yrs > 110 | 
        (is.na(deathdate) & (
          !is.na(death_city) | !is.na(death_state) | !is.na(death_country) |
            !is.na(cemetery) | !is.na(ceme_city) | !is.na(ceme_state) |
            !is.na(ceme_country) | !is.na(ceme_note))), 
      NA_real_, 
      life_yrs
    ),
    career_yrs = time_length(
      interval(play_debut, play_lastgame), 
      "years"
    ),
    play_lastgame_life_yrs = time_length(
      interval(birthdate, play_lastgame), 
      "years"
    ),
    # Event status (1 = event, 0 = censor, NA = don't know)
    life_status = case_when(
      is.na(life_yrs) ~ NA_integer_,  # don't know
      !is.na(deathdate) ~ 1L,  # died
      TRUE ~ 0L  # alive
    ),
    # There is no retirement date. Assume its the last game played for cases
    # where the player has been out of the game for over 2 years.
    yrs_since_lastgame = time_length(
      interval(play_lastgame, bio_last_updated), 
      "years"
    ),
    career_status = case_when(
      is.na(career_yrs) ~ NA_integer_,  # don't know
      yrs_since_lastgame > 2 ~ 1L,  # no longer playing
      TRUE ~ 0L # still playing
    ) 
  ) %>%
  select(-c(starts_with("death_"), starts_with("ceme"), yrs_since_lastgame))
```

One more data change. `height` is in format FT-IN. Parse it to calculate the height in inches.

```{r}
bio_4 <- bio_3 %>%
  mutate(
    tmp_height = map(height, ~str_split(.x, "-", n = 2) %>% pluck(1)),
    tmp_height_ft = map_dbl(
      tmp_height, 
      ~if_else(is.null(.x), "", pluck(.x, 1)) %>% as.numeric()
    ),
    tmp_height_in = map_dbl(
      tmp_height, 
      ~if_else(is.null(.x), "", pluck(.x, 2)) %>% as.numeric()
    ),
    tmp_height_ft_2 = if_else(tmp_height_ft <= 7, tmp_height_ft, NA_real_),
    height_in = tmp_height_ft_2 * 12 + tmp_height_in
  ) %>% 
  select(-c(starts_with("tmp_"), height))
```

Anything else interesting? Maybe body-mass index from the height and weight.

```{r}
bio_5 <- bio_4 %>%
  mutate(
    bmi = weight / height_in^2 * 703
  )
```

Final look at the data.

```{r}
bio <- bio_5
skimr::skim(bio)
```

## Exploration

This section is primarily univariate analyses to get a feel for the data and to establish expectations for the later survival models.

### Height and Weight

According to a study [published in the journal of eLife](https://elifesciences.org/articles/13410), being taller is associated with enhanced longevity and with higher education, earnings, and social position. The average man born in 1896 grew to 5' 7" (67"). The average for 1996 was 5' 10" (70"). The Retrosheet data indicates baseball players are quite a bit taller (Table 1, Figs. 1-2). The average height for players born in 1896 was 5' 11" (71") and 6' 2" (74") for players born in 1996. Players are getting heavier too. Today's athlete is over 35 pounds heavier than those born 100 years ago. 

```{r}
bio %>% 
  mutate(birthyear = year(birthdate)) %>%
  filter(birthyear %in% c(1850, 1896, 1996)) %>%
  gtsummary::tbl_summary(
    by = birthyear,
    statistic = list(all_continuous() ~ "{mean}, {median} ({p25}, {p75})"),
    digits = list(height_in ~ 0, weight ~ 0, bmi ~ 1),
    include = c("height_in", "weight", "bmi")
  ) %>% modify_caption("Table 1. MLB pLayer size for select birth years.")
```

```{r}
p1 <- bio %>%
  filter(!is.na(birthdate) & !is.na(height_in)) %>%
  mutate(yr = lubridate::year(birthdate)) %>%
  ggplot(aes(x = yr, y = height_in)) +
  geom_point(alpha = .2, color = mlb_palette["navy_blue"]) +
  geom_smooth(method = "lm", formula = "y~x", color = "goldenrod") +
  labs(
    title = "Fig 1. MLB Players Are Getting Taller.",
    subtitle = "Height by year of birth.",
    y = "height (inches)", x = NULL
  ) +
  theme_light() +
  mlb_theme 
p1
```

```{r}
p2 <- bio %>%
  filter(!is.na(birthdate) & !is.na(weight)) %>%
  mutate(yr = lubridate::year(birthdate)) %>%
  ggplot(aes(x = yr, y = weight)) +
  geom_point(alpha = .2, color = mlb_palette["navy_blue"]) +
  geom_smooth(method = "lm", formula = "y~x", color = "goldenrod") +
  labs(
    title = "Fig 2. MLB Players Are Getting Heavier.",
    subtitle = "Weight by year of birth.",
    y = "weight (pounds)", x = NULL
  ) +
  theme_light() +
  mlb_theme 
p2
```

Height and weight can be combined into a single measure of mass, the BMI.

$$
BMI = \frac{weight}{height^2} \times 703
$$

```{r}
p3 <- bio %>%
  filter(!is.na(birthdate) & !is.na(bmi)) %>%
  mutate(yr = lubridate::year(birthdate)) %>%
  ggplot(aes(x = yr, y = bmi)) +
  geom_point(alpha = .2, color = mlb_palette["navy_blue"]) +
  geom_smooth(method = "lm", formula = "y~x", color = "goldenrod") +
  labs(
    title = "Fig 3. MLB Players Are Getting More Massive.",
    subtitle = "Body Mass Index (BMI) by year of birth.",
    y = "bmi", x = NULL
  ) +
  theme_light() +
  mlb_theme 
p3
```

Does the Retrosheet data confirm the eLife study? No. Correcting for birth year (a secular trend of improving health), each inch of height is associated with 2 *less* months (.17 yrs) of life.

```{r}
bio %>% 
  filter(life_status == 1) %>% 
  mutate(birth_year = year(birthdate)) %>%
  lm(life_yrs ~ height_in + birth_year, data = .) %>%
  gtsummary::tbl_regression()
```

### Handedness {.tabset}

According to [this Psychology Today blog](https://www.psychologytoday.com/us/blog/the-asymmetric-brain/201908/8-new-scientific-findings-about-left-handedness)), about 9% of people are left-handed. However among Major League baseball players the prevalence is much higher. Fig. 4a shows handedness by birth year. The data is a little messy because data is less available for early years, and in the very latest years the proportions are noisy due to low sample size (not very many players today were born after 2000).

Batting left-handed may be as much a product of strategy, or chance as it is a product of handedness. Throwing almost certainly is a product of handedness. You can see that in the number of ambidextrous pitchers - there are amost none. The Psychology Today article suggests cultural norms play a role in handedness. Fig 4b does seem to show an increasing trend in the proportion of left-handed pitchers up until 1950.

```{r}
plot_p4 <- function(dat) {
  dat %>%
    filter(!is.na(birthdate)) %>%
    mutate(birth_year = year(birthdate)) %>%
    pivot_longer(cols = c(bats, throws)) %>%
    mutate(value = factor(value, levels = c("L", "R", "B"), labels = c("Left", "Right", "Both"))) %>%
    group_by(birth_year, name, value) %>%
    summarize(.groups = "drop_last", n = n()) %>%
    mutate(pct = n / sum(n)) %>%
    ungroup() %>%
    ggplot(aes(x = birth_year, y = pct, color = value)) +
    geom_line() +
    scale_y_continuous(labels = percent_format(1)) +
    facet_wrap(facets = vars(name)) +
    theme_light() +
    mlb_theme +
    theme(
      strip.background = element_rect(fill = "#E8D8C4", color = "#C49D6E"),
      strip.text = element_text(color = "#231F20")) +
    labs(
      x = "Birth Year", y = NULL, color = NULL
    )
}
```
#### Fig 4a

```{r}
p4a <- bio %>% plot_p4()
p4a +
  scale_color_manual(values = c("Left" = "#F4793E", "Right" = "#231F20", "Both" = "#C49D6E", "N/A" = "#E8D8C4")) +
  labs(title = "Fig 4a. Handeness prevalance (raw).")
```

#### Fig 4b

```{r}
p4b <- bio %>% 
  filter(!is.na(bats) & !is.na(throws)) %>% 
  plot_p4() +
  scale_color_manual(values = c("Left" = "#F4793E", "Right" = "#231F20", "Both" = "#C49D6E")) +
  labs(title = "Fig 4b. Handeness prevalance (cleaned).")
p4b
```

### Birthplace {.tabset}

MLB has a relatively high concentration of ballplayers of Latin descent. Many were born outside the US. Of ballplayers born in the US, most are naturally from high population states. Adding population data to the data set will normalize the player counts for comparison. 

```{r message=FALSE, warning=FALSE}
# Help from https://r-spatial.org/r/2018/10/25/ggplot2-sf.html.

library(sf)
library(rnaturalearth)
library(rnaturalearthdata)
library(ggspatial)

# Geo data.
world_dat <- ne_countries(scale = "medium", returnclass = "sf")
state_dat <- ne_states(country = "United States of America", returnclass = "sf")

# Make X, Y coordinates for country and state labels.
world_centroid <- world_dat %>% 
  st_make_valid() %>% 
  st_centroid() %>% 
  st_coordinates()
state_centroid <- state_dat %>% 
  st_make_valid() %>% 
  st_centroid() %>% 
  st_coordinates()

# Modify country names in bio to match those in world_dat.
bio_country <- bio %>% 
  mutate(
    birth_country = case_when(
      birth_country == "Dominican Republic" ~ "Dominican Rep.",
      birth_country == "Curacao" ~ "Curaçao",
      birth_country == "England" ~ "United Kingdom",
      birth_country == "Canal Zone" ~ "Panama",
      birth_country == "USA" ~ "United States",
      birth_country == "South Korea" ~ "Korea",
      birth_country == "South Vietnam" ~ "Vietnam",
      birth_country == "Virgin Islands" ~ "U.S. Virgin Is.",
      birth_country == "West Germany" ~ "Germany",
      birth_country == "Austria-Hungary" ~ "Austria",
      birth_country == "British Honduras" ~ "Honduras",
      birth_country == "Czechoslovakia" ~ "Czech Rep.",
      birth_country == "Northern Ireland" ~ "Ireland",
      birth_country == "Dominican Republic" ~ "Dominican Rep.",
      TRUE ~ birth_country
    )
  )

# Current players per capita
bio_country_n <- bio_country %>% 
  filter(career_status == 0) %>%
  count(birth_country) %>%
  left_join(world_dat %>% select(name, pop_est), 
            by = c("birth_country" = "name")) %>%
  mutate(n_per_capita = n / pop_est)

census_pop <- readr::read_csv(
  "https://www2.census.gov/programs-surveys/popest/datasets/2020-2021/state/totals/NST-EST2021-alldata.csv"
)

bio_state_n <- bio %>%
  filter(career_status == 0 & birth_country == "USA") %>%
  count(birth_state) %>%
  inner_join(census_pop, by = c("birth_state" = "NAME")) %>%
  mutate(n_per_capita = n / POPESTIMATE2021)
```

#### International

Today, 28% of MLB players were born outside the US. The Dominican Republic is the largest source of talent (`r bio_country_n %>% filter(birth_country == "Dominican Rep.") %>% pull(n)` players), and has a stunning per capita contribution of `r bio_country_n %>% filter(birth_country == "Dominican Rep.") %>% mutate(n_per_capita = n_per_capita * 1E6) %>% pull(n_per_capita) %>% comma(.1)` players per million of population. The tiny island of Curaçao, population 142,000, boasts `r bio_country_n %>% filter(birth_country == "Dominican Rep.") %>% pull(n)` players for a whopping per capita contribution of `r bio_country_n %>% filter(birth_country == "	
Curaçao") %>% mutate(n_per_capita = n_per_capita * 1E6) %>% pull(n_per_capita) %>% comma(.1)` players per million.

```{r message=FALSE, warning=FALSE}
world_dat %>%
  cbind(world_centroid) %>%
  left_join(bio_country_n, by = c("name" = "birth_country")) %>%
  replace_na(list(n_per_capita = 0)) %>% 
  mutate(
    lbl = if_else(name %in% c("United States", "Mexico", "Dominican Rep.", "Cuba", "Venezuela", "Curaçao"),
                  glue("{abbrev}\n{comma(n_per_capita * 1E6, .1)}"),
                  glue(""))
  ) %>% 
  ggplot() +
  geom_sf(aes(fill = n_per_capita), show.legend = FALSE) +
  geom_text(aes(x = X, y = Y, label = lbl),
    color = "black", 
    # fontface = "bold", 
    check_overlap = FALSE,
    # family = "TR Plaza",
    size = 3.0
  ) +
  scale_fill_gradient(low = "white", high = "firebrick") +
  annotation_scale(location = "bl", width_hint = 0.25) +
  annotation_north_arrow(location = "bl", which_north = "true", 
                         pad_x = unit(0.5, "in"), pad_y = unit(0.5, "in"),
                         style = north_arrow_fancy_orienteering) +
  coord_sf(xlim = c(-130, -45), ylim = c(0, 50)) +
  theme_light() +
  mlb_theme +
  labs(
    title = "Fig 5a. The Dominican Republic is a baseball powerhouse.", 
    subtitle = "Count of Major Leaguers per Million of Population.",
    x = NULL, y = NULL, fill = NULL
  ) 
```

#### Domestic

California (`r bio_state_n %>% filter(birth_state == "California") %>% pull(n)`) and Florida (`r bio_state_n %>% filter(birth_state == "Florida") %>% pull(n)`) are not only the highest overall producers of talent, they are also the highest per-capita producers (along with Georgia).

```{r warning=FALSE, message=FALSE}
state_dat %>%
  cbind(state_centroid) %>%
  left_join(bio_state_n, by = c("name" = "birth_state")) %>%
  replace_na(list(n_per_capita = 0)) %>%
  mutate(lbl = comma(n_per_capita * 1E6, .1)) %>%
  ggplot() +
  geom_sf(aes(fill = n_per_capita), show.legend = FALSE) +
  geom_text(aes(x = X, y = Y, label = lbl),
    color = "black",
    # fontface = "bold",
    check_overlap = TRUE,
    # family = "TR Plaza",
    size = 3.0
  ) +
  scale_fill_gradient(low = "white", high = "firebrick") +
  annotation_scale(location = "bl", width_hint = 0.25) +
  annotation_north_arrow(location = "bl", which_north = "true", 
                         pad_x = unit(0.5, "in"), pad_y = unit(0.5, "in"),
                         style = north_arrow_fancy_orienteering) +
  coord_sf(xlim = c(-130, -60), ylim = c(20, 50)) +
  theme_light() +
  mlb_theme +
  labs(
    title = "Fig 5b. Top Domestic Sources of Talent.", 
    subtitle = "Count of Major Leaguers per Million of Population.",
    x = NULL, y = NULL, fill = NULL
  ) 
```

### {}

Major League baseball was not an international organization until after WWII (Fig 6). Interesting that the earliest ballplayers were often born in Europe.

```{r}
bio_country %>%
  left_join(world_dat, by = c("birth_country" = "name")) %>%
  mutate(debut_year = year(play_debut)) %>%
  count(debut_year, region_wb) %>%
  group_by(debut_year) %>%
  mutate(pct = n / sum(n)) %>%
  ungroup() %>%
  filter(!is.na(region_wb)) %>%
  mutate(
    region_wb = fct_relevel(
      region_wb, 
      "North America", "Latin America & Caribbean", "East Asia & Pacific"
    ),
    region_wb = fct_collapse(
      region_wb,
      `North America` = "North America",
      `Latin America & Caribbean` = "Latin America & Caribbean", 
      `East Asia & Pacific` = "East Asia & Pacific",
      `Europe & Central Asia` = "Europe & Central Asia",
      other_level = "Other"
    )
  ) %>%
  ggplot(aes(x = debut_year)) +
  geom_line(aes(y = pct, color = region_wb)) +
  labs(
    title = "MLB became more international after WWII.",
    subtitle = "Players debuting by year.",
    x = NULL, y = NULL, color = NULL
  ) +
  scale_y_continuous(labels = percent_format(1)) +
  theme_light() +
  mlb_theme
```


```{r}
# Switch-hitters are more likely to join the Hall. Next best is batting left.
bio %>%
  tabyl(bats, hof) %>%
  adorn_percentages(denominator = "row") %>%
  adorn_pct_formatting() %>%
  adorn_ns() %>%
  kableExtra::kbl()

# ...but it's better to throw right-handed
bio %>%
  tabyl(throws, hof) %>%
  adorn_percentages(denominator = "row") %>%
  adorn_pct_formatting() %>%
  adorn_ns() %>%
  kableExtra::kbl()
```

### Hall of Famers

## KM

km_fit <- survfit(Surv(life_yrs))


## Reference

* [Baseball Data Science](https://www.baseballdatascience.com/survival-analysis-how-long-do-careers-last/)

* [NIH](https://www.ncbi.nlm.nih.gov/pmc/articles/PMC2394262/)