---
title: "Reproducing the Abel and Kruger Study"
author: "Michael Foley"
date: "`r Sys.Date()`"
output: 
  html_document:
    css: "style.css"
    theme: flatly
    toc: true
    toc_float: true
    highlight: haddock
    code_folding: show
knit: (function(inputFile, encoding) {
  rmarkdown::render(inputFile, encoding = encoding, output_dir = "../docs") })
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Abel and Kruger ([2005](#Abel2005)) use professional baseball player data to evaluate the role of awarded achievement within the sport on longevity. The authors hypothesize that players inducted into the Baseball Hall of Fame have longer longevities than players of the same age who are not inducted. This workbook reproduces the results from the same source data and methods.

```{r include=FALSE}
# library(Lahman)  # has latest data
library(tidyverse)
library(lubridate)
library(scales)
```

# Data

The authors used the 2002 Lahman database, Version 5.0 ([2002](#Lahman2002)). I downloaded the Access database and exported the **HallOfFame** and **Master** tables to csv.^[There is also a zipped csv archive, but it lacks column headers.]

```{r warning=FALSE, message=FALSE}
Master <- read_csv("../data/lahman_50-2k/Master.csv")
HallOfFame <- read_csv("../data/lahman_50-2k/HallOfFame.csv")
```

`Master` includes non-players (executives, announcers, and umpires) who were inducted into the Hall of Fame. Actual players are identified by a non-empty `playerID` column. Of the `r comma(nrow(Master))` rows in the data set, `r comma(n_distinct(Master$playerID))` are players.^[To confirm my understanding, I compared the `playerID` values from the **Batting**, **Fielding**, and **Pitching** tables. The union of values is the same as those in `Master`.]

```{r include=FALSE, eval=FALSE}
Batting <- read_csv("../data/lahman_50-2k/Batting.csv")
Fielding <- read_csv("../data/lahman_50-2k/Fielding.csv")
Pitching <- read_csv("../data/lahman_50-2k/Pitching.csv")

player_ids <- c(Batting$playerID, Fielding$playerID, Pitching$playerID) %>% unique()

# Yes, playerID indicates an actual player.
Master %>% filter(!is.na(playerID)) %>% nrow()
Master %>% filter(playerID %in% player_ids) %>% nrow()
Master %>% filter(!playerID %in% player_ids & !is.na(playerID)) %>% nrow()
```

The *Results* section of the article summarizes the row counts of the analysis data set derived from these tables (page 961).

> One hundred and forty-three (143) players in the Hall of Fame had been inducted into the Baseball Hall of Fame while still alive as of 2002. These were age-matched with 3,430 players. The mean age of the Hall of Famers at time of induction
was 57.5 years (SD $\pm$ 12.4). A total of 1,695 of these players had died by the end of 2002.

The code chunk below identifies the 143 Hall of Famers as object `hall_of_famers`. The criteria is in the filter: 1) inducted into the Hall, 2) as a player, 3) through 2002, and 4) while alive.

```{r}
hall_of_famers <- HallOfFame %>% 
  inner_join(Master, by = "hofID") %>%
  filter(
    inducted == "Y",
    # Inducted as a player _and_ has a playerID identifier (need both!)
    category == "Player", !is.na(playerID),
    yearid <= 2002,
    coalesce(deathYear, 9999) > yearid
  ) %>% 
  mutate(age_at_induction = yearid - birthYear) %>%
  select(playerID, inductionYear = yearid, birthYear, age_at_induction)
```

Does this match the article? Not quite. I get 144 rows^[I get 143 rows if I exclude the 2002 induction year, but the age mean and SD are still off.], and the mean $\pm$ SD age at induction is slightly off.

```{r collapse=TRUE}
# Does not match 143 in article.
nrow(hall_of_famers)

# Does not match 57.5 in article.
mean(hall_of_famers$age_at_induction)

# Does not match 12.4 in article.
sd(hall_of_famers$age_at_induction)
```

The *Methods and Materials* section of the article elaborates the age-matching criteria (page 960).

> [age-matched players] were alive at the time of induction when induction occurred for their case-matched cohort. Only players were included in our analysis (executives, announcers, umpires were not included; managers were included who had also been players).

The code chunk below identifies every player who shares the same age as a Hall of Famer _and_ was alive at that Hall of Famer's induction year. Note that a player may have been alive for one Hall of Famer's induction, but not for the later induction of another Hall of Famer with the same birth year. 

```{r}
age_matched_cohort <- Master %>%
  inner_join(hall_of_famers, by = "birthYear", suffix = c("", ".hof")) %>%
  filter(
    !is.na(playerID),
    coalesce(deathYear, 9999) > inductionYear
  ) %>%
  anti_join(hall_of_famers, by = "playerID") %>%
  select(playerID, birthYear, deathYear, nameFirst, nameLast, hofID) %>%
  # A player can match multiple Hall of Famers
  unique()
```

Does this match the article? Not even close. `age_matched_cohort` should have 3,430 players. Instead, it has `r scales::comma(nrow(age_matched_cohort))` players. I tried defining age-match as sharing both birth year and birth month, but that yielded a result set that was way too *small*.

```{r collapse=TRUE}
nrow(age_matched_cohort)
```

## What went wrong?

What went wrong? Let's audit 20 random players of `age_matched_cohort`. Since `age_matched_cohort` has so many more players than it should have, at least one of these players should be unjustified.

```{r}
set.seed(1234)
random_rownums <- runif(20, 1, nrow(age_matched_cohort)) %>% as.integer()

matches <- age_matched_cohort[random_rownums, ] %>%
  inner_join(hall_of_famers, by = "birthYear", suffix = c(".cohort", ".hof")) %>%
  filter(coalesce(deathYear, 9999) > inductionYear,
         playerID.cohort != playerID.hof) %>%
  select(playerID.cohort, nameFirst, nameLast, birthYear, deathYear,
         playerID.hof, inductionYear)
```

A player is age-matched if 1) there was a Hall of Famer inducted while alive who shares the same birth year, and 2) they were alive during the induction year too. Are these players correctly age-matched correctly? Yes, all 20 are age-matched.

```{r collapse=TRUE}
matches %>% select(playerID.cohort) %>% unique() %>% nrow()
```

Let's look at some of the matches. Ethan Allen was still alive when Red Ruffing was inducted in 1967. Both were born in 1904. Art Bues was alive when Tris Speaker was inducted in 1937, and both were born in 1888. Douglas Whammy shared the same birth year as three Hall of Famers inducted while still alive. Douglas Whammy was still alive as of 2002.

```{r}
matches %>% rename(inducted = inductionYear) %>% arrange(playerID.cohort)
```

Why is my data set so much longer than Abel and Kruger's - what am I missing here?

# References

<a id="Abel2005"></a>Ernest L. Abel & Michael L. Kruger (2005) The Longevity of Baseball Hall of Famers Compared to Other Players, Death Studies, 29:10, 959-963, DOI: [10.1080/07481180500299493](https://doi.org/10.1080/07481180500299493). [PDF](http://www.med.mcgill.ca/epidemiology/hanley/bios601/CandHchapter06/baseball_players.pdf).

<a id="Lahman2002"></a>[Sean Lahman's Baseball Database](https://www.seanlahman.com/baseball-archive/statistics/).
